---
title: "Statlearninghw1"
author: "Sam Reade, Ava Exelbirt"
date: "2024-11-08"
output: html_document 
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statistical Learning Homework 1:

```{r libraries}
#install libraries needed 
#install.packages("ggcorrplot")
#install.packages("psych")
#install.packages("factoextra")
#install.packages("pheatmap")
#install.packages("dbscan")
library(dbscan)
library(pheatmap)
library(cluster)
library(factoextra)
library(psych)
library(ggcorrplot)
library(dplyr)
library(tidyr)
library(ggplot2)
library(dendextend)
library(reshape2)
library(mclust)
```

# Introduction:

Using a data set from TidyTueday that contains information about rent
prices for houses in the Bay Area of California. The data set contains
information regarding date of rental, location of the house (like city
and county for example), information regarding size of house, and price
of rent. The original data set Rent had 200,769 observations among 17
variables. After data cleanup and processing, the variables we chose to
include in our model were year, neighborhood, city, county, price, beds,
baths, soft, and title. Year is the year the house was on the market,
neighborhood, city, and county outline the area the house is located in
the Bay Area, price is the rental price is US dollars per one month of
rent, beds is the amount of bedrooms the house has, baths is the amount
of bathrooms the house has, sqft is the amount of sqft the house has,
and title is the title of the rental listing posting. We chose these
variables because they are the most representative of houses in the Bay
Area in relation to our research question. 

Using the rent data set we aim to explore and understand the
relationships among various housing characteristics, such as price,
square footage, number of bedrooms, and bathrooms, using Principal
Component Analysis (PCA), Factor Analysis, and Clustering for houses in
Bay Area, California. Some relationships we can predict are: predicted
price by square footage, predicting price by county, and predicting
county by square footage. Overall, we can investigate the relationships
among variables: we can predict house prices based on square footage,
number of bedrooms, and bathrooms. We can analyze how price-per-unit
metrics (e.g., price per square foot) vary by cluster. We can also
explore potential regional or neighborhood-level trends. Overall, we can
use PCA, FA, and the clustering results to understand how different
house characteristics correlate with pricing tiers or market
segmentation.

```{r get_df}
#Load data set from Tidy Tuesday
Rent_df = tidytuesdayR::tt_load('2022-07-05')
```

```{r}
#Organize and rename datasets
Rent = Rent_df$rent
Permits = Rent_df$sf_permits
Construction = Rent_df$new_construction
```

# DATA PREPROCESSING:

```{r}
head(Rent)
dim(Rent)
```

## Data Cleanup:

We can first check the number of duplicate rows as a cautionary check.

```{r}
# Check number of Duplicate Rows: 
num_duplicates = sum(duplicated(Rent))
print(num_duplicates)
```

## Checking unique values of categorical features:

We will print the distinct values represented in the data set for each
categorical variable and year.

### Year values:

```{r}
unique_year = unique(as.factor(Rent$year))
print(unique_year)

years_count = table(Rent$year)
print(years_count)
```

### Neighborhood values:

```{r}
unique_nhood = unique(as.factor(Rent$nhood))
print(unique_nhood)

nhood_count = table(Rent$nhood)
print(nhood_count)

# 167 nhoods
```

### City values:

```{r}
unique_city = unique(as.factor(Rent$city))
print(unique_city)

city_count = table(Rent$city)
print(city_count)

# 104 cities
```

### County values:

```{r}
unique_county = unique(as.factor(Rent$county))
print(unique_county)

county_count = table(Rent$county)
print(county_count)

# 10 counties
```

## Removing unnecessary columns:

We will only use the columns we deem appropriate for our analysis: year,
neighborhood, city, county, price, beds, baths, soft, and title.
Therefore, we will delete room_in_apt, lat, lon, address, descr, and
details from the dataframe.

```{r}
Rent = Rent %>% select(-room_in_apt, -lat, -lon, -address, -descr, -details)
head(Rent)
```

## Checking count of rows with at least 1 NA value:

```{r}
count_na_rows = sum(apply(Rent, 1, function(x) any(is.na(x))))
print(count_na_rows)
```

## Checking count of NA value in each column:

```{r}
na_count_year = sum(is.na(Rent$year))
print(paste("NA count year:", na_count_year))

na_count_nhood = sum(is.na(Rent$nhood))
print(paste("NA count nhood:", na_count_nhood))

na_count_city = sum(is.na(Rent$city))
print(paste("NA count city:", na_count_city))

na_count_county = sum(is.na(Rent$county))
print(paste("NA count county:", na_count_county))

na_count_price = sum(is.na(Rent$price))
print(paste("NA count price:", na_count_price))

na_count_beds = sum(is.na(Rent$beds))
print(paste("NA count beds:", na_count_beds))

na_count_baths = sum(is.na(Rent$baths))
print(paste("NA count baths:", na_count_baths))

na_count_sqft = sum(is.na(Rent$sqft))
print(paste("NA count sqft:", na_count_sqft))

na_count_title = sum(is.na(Rent$title))
print(paste("NA count title:", na_count_title))

```

## Removing rows with NAs for title, county, beds, baths, and sqft

```{r}
#cleanup data to include rows with no NA values
Rent_cleaned <- Rent |>
  filter(!is.na(title), !is.na(county), !is.na(beds), !is.na(baths), !is.na(sqft))

Rent_cleaned
```

## Checking for complete rows (no NAs):

```{r}
complete_rows = sum(complete.cases(Rent))

print(complete_rows)
```

## Checking for Outliers:

```{r}

detective = function(column) {

  Q1 = quantile(column, 0.25, na.rm = TRUE)
  Q3 = quantile(column, 0.75, na.rm = TRUE)
  IQR = Q3 - Q1
  
  lower_bound = Q1 - 1.5 * IQR
  upper_bound = Q3 + 1.5 * IQR
  
  outliers = column < lower_bound | column > upper_bound
  
  return(outliers)
}

price_outliers = detective(Rent_cleaned$price)
num_outliers_price = sum(price_outliers)
print(paste("Price outlier count:", num_outliers_price))

bed_outliers = detective(Rent_cleaned$beds)
num_outliers_bed = sum(bed_outliers)
print(paste("Bed outlier count:", num_outliers_bed))

sqft_outliers = detective(Rent_cleaned$sqft)
num_outliers_sqft = sum(sqft_outliers)
print(paste("sqft outlier count:", num_outliers_sqft))

bath_outliers = detective(Rent_cleaned$baths)
num_outliers_bath = sum(bath_outliers)
print(paste("Bath outlier count:", num_outliers_bath))
```

## Visualizing Outliers:

```{r}
boxplot(Rent_cleaned$price, 
        main = "Box Plot of Rent Prices", 
        ylab = "Price", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$baths, 
        main = "Box Plot of baths", 
        ylab = "baths", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$sqft, 
        main = "Box Plot of sqft", 
        ylab = "sqft", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$beds, 
        main = "Box Plot of beds", 
        ylab = "beds", 
        col = "lightblue",
        outline = TRUE)  


```

## Remove Outliers:

```{r}
#only keep rows with no outliers present
Rent_cleaned <- Rent_cleaned |>
  filter(
    !detective(price),
    !detective(beds),
    !detective(baths),
    !detective(sqft)
  )
```

## Visualizations of Variable Distributions:

### Histogram of Price:

```{r}
ggplot(Rent_cleaned, aes(x = price)) +
  geom_histogram(binwidth = 250, color = "black", fill = "skyblue") +
  labs(title = "Histogram of Price", x = "Price", y = "Count") +
  theme_minimal() +
  coord_cartesian(xlim = c(0, 20000))
```

### Histogram of sqft:

```{r}
ggplot(Rent_cleaned, aes(x = sqft)) +
  geom_histogram(binwidth = 100, color = "black", fill = "skyblue") +
  labs(title = "Histogram of Price", x = "sqft", y = "Count") +
  theme_minimal() + 
  coord_cartesian(xlim = c(0, 10000))
```

### Bar Plot of beds:

```{r}
ggplot(Rent_cleaned, aes(x = beds)) +
  geom_bar(color = "black", fill = "skyblue") +
  labs(title = "Bar Plot of beds", x = "Number of beds", y = "Count") +
  theme_minimal()
```

### Bar Plot of baths:

```{r}
ggplot(Rent_cleaned, aes(x = baths)) +
  geom_bar(color = "black", fill = "skyblue") +
  labs(title = "Bar Plot of baths", x = "Number of baths", y = "Count") +
  theme_minimal()
```

### Bar Plot for year:

```{r}
ggplot(Rent_cleaned, aes(x = factor(year))) +  
  geom_bar(color = "black", fill = "skyblue") +  
  labs(title = "Bar Plot of Year", x = "Year", y = "Count") +  
  theme_minimal()
```

# Feature Engineering:

Create 3 new variables: price_per_sqft, price_per_bed, and
price_per_bath which can be calculated by taking the rent price divided
by sqft, bed, bath, respectively.

```{r}
#create a new data set that adds the new feature engineered variables and does not include title. 
Rent_num = Rent_cleaned

price_per_sqft = Rent_cleaned$price / Rent_cleaned$sqft 
price_per_bed = Rent_cleaned$price / Rent_cleaned$beds
price_per_bath = Rent_cleaned$price / Rent_cleaned$baths

Rent_num$price_per_sqft = price_per_sqft
Rent_num$price_per_bed = price_per_bed
Rent_num$price_per_bath = price_per_bath

Rent_num = Rent_num |> select(-title)

Rent_num
```

## Visualizations of the New Features

```{r}
ggplot(Rent_num, aes(x = price_per_sqft)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightgreen") +
  labs(title = "Histogram of Price per Sqft", x = "Price per Sqft", y = "Count") +
  theme_minimal()

ggplot(Rent_num, aes(x = price_per_bed)) +
  geom_histogram(binwidth = 100, color = "black", fill = "lightcoral") +
  labs(title = "Histogram of Price per Bed", x = "Price per Bed", y = "Count") +
  theme_minimal()

ggplot(Rent_num, aes(x = price_per_bath)) +
  geom_histogram(binwidth = 200, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Price per Bed", x = "Price per Bed", y = "Count") +
  theme_minimal()
```

# EDA Visualizations for Relationships Among Variables:

```{r}
ggplot(Rent_num, aes(x = sqft, y = price)) + 
  geom_point(aes(color = price), alpha = 0.6) + 
  scale_color_gradient(low = "blue", high = "red") + 
  labs(
    title = "Price vs. Square Footage",
    x = "Square Footage",
    y = "Price"
  ) + 
  theme_minimal() + 
  theme(legend.position = "none")


ggplot(Rent_num, aes(x = county, y = price)) +
  geom_boxplot(fill = "lightblue", color = "blue", outlier.colour = "red") +
  labs(title = "Price Distribution by county", 
       x = "Neighborhood", 
       y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(Rent_num, aes(x = sqft, y = price_per_sqft)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(title = "Price per Square Foot vs. Square Footage",
       x = "Square Footage",
       y = "Price per Square Foot") +
  theme_minimal()


ggplot(Rent_num, aes(x = baths, y = price_per_bath)) +
  geom_point(color = "green", alpha = 0.6) +
  labs(title = "Price per Bath vs. Number of Baths",
       x = "Number of Baths",
       y = "Price per Bath") +
  theme_minimal()
```

## Scale Numeric Features

Standardize/normalize numeric features in dataset.

```{r}
#Scale the numeric variables in the data set so thet are processed for PCA
numeric_cols <- Rent_num |> select(where(is.numeric)) |> names()
Rent_num[numeric_cols] <- scale(Rent_num[numeric_cols])
```

## Correlation Analysis with the new features:

```{r}
#calculate the correlation matrix for the variables in our cleaned data set 
numeric_complete_rent2 <- Rent_num |> select(price_per_bath, price_per_bed, price_per_sqft, sqft, baths, beds, price)

cor_matrix2 <- cor(numeric_complete_rent2, use = "pairwise.complete.obs")

ggcorrplot(cor_matrix2, lab = TRUE)
```

## Data Set/Extra Cleanup for PCA:

Create a new data set Rent_Analysis that will be used for PCA. It
includes all numeric features and has extra checks to make sure there
are no NA or infinite values.

```{r}
#Clean the data by handling NA and infinite values. Extra cautionary check before going into PCA
#Replace NA values with the column mean, and handle Infinite values if they are there
Rent_analysis <- Rent_num |>
  select(price, sqft, beds, baths, price_per_sqft, price_per_bed, price_per_bath)

Rent_analysis <- Rent_analysis |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>
  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .))) %>%
  drop_na() 
```

# PCA ANALYSIS:

## Perform PCA

First standardize the data, then perform PCA, and view the PCA loadings.
The PCA loadings (the relationship between original variables and
principal components). This will help understand how each feature
contributes to each principal component

```{r PCA}
#make sure data is clean with no NA and no 0 variance observations. 
Rent_PCA <- Rent_analysis |>
  select(where(~ !all(is.na(.))))

Rent_PCA <- Rent_PCA |>
  select(where(~ var(.) != 0))

#Rent_PCA_scaled <- scale(Rent_PCA) #data set was scaled above in Rent_analysis

pca_result <- prcomp(Rent_PCA, center = TRUE, scale. = TRUE)

# Scree Plot to explain variance explained by each PC
fviz_screeplot(pca_result, addlabels = TRUE, title = "Scree Plot")

summary(pca_result)

print(pca_result$rotation)
```

## Proportion of Variance for each PC

This tells you how much of the variability in the data is explained by
each component

```{r}
proportion_variance <- summary(pca_result)$importance[2, ]
print(proportion_variance)
```

# PC Visualizations.

First, print the PC results:

```{r}
#print data from first two PC results
Rent_PCA_df = data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2])
Rent_PCA_df
```

## First two PC Visuals

Visualize the first two PC loadings in relation to each other and
individually.

```{r}
#visualize the first two PC loadings
#PC1 vs PC2 plot
ggplot(Rent_PCA_df, aes(x = PC1, y = PC2)) +
  geom_point(color = "blue", alpha = 0.6) +  
  theme_minimal() +
  labs(title = "PCA: First vs. Second Principal Component", 
       x = "Principal Component 1", y = "Principal Component 2")

#loadings for PC1 visual
barplot(pca_result$rotation[,1], las = 2, col = "darkblue", main = "Loadings for PC1")
#loadings for PC2 visual
barplot(pca_result$rotation[,2], las = 2, col = "darkblue", main = "Loadings for PC2")
```

## Visualize the Variance of PCs.

```{r}
variance_explained <- summary(pca_result)$importance[2, ]
ggplot(data.frame(PC = 1:length(variance_explained), Variance_Explained = variance_explained), 
       aes(x = factor(PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Variance Explained by Each Principal Component", x = "Principal Component", y = "Variance Explained")
```

## Contribution of Variables to PCs

This shows how much each variable is contributing to the two PCs. The
bar heights indicate how much each variable contributes to the variance
captured by PC1 or PC2. A high contribution means that the variable is
important for explaining the variation along PC1 or PC2. For PC1,
price_per_bed is the most important for explaining the variance captured
by PC1, followed by price_per_sqft. For PC2, price is the most important
for explaining the variance captured by PC2, with all other variables at
least 20% lower in explaining variance.

```{r}
fviz_contrib(pca_result, choice = "var", axes = 1, title = "Contribution of Variables to PC1")
fviz_contrib(pca_result, choice = "var", axes = 2, title = "Contribution of Variables to PC2")

```

## PCA Bipolot

Use the Biplot to visualize both the **scores** (the transformed data
points) and the **loadings** (the contributions of the original
variables) on the same plot. It helps understand how the original
variables relate to each principal component and how the data points are
distributed in the reduced-dimensional space.

```{r}
fviz_pca_biplot(pca_result, repel = TRUE, title = "PCA Biplot")
                
```

## Create dataframe with results

The principal components (scores) are stored in pca_result\$x, which we
convert into a dataframe.

```{r}
pca_data <- data.frame(pca_result$x) 
```

## Individual Component Contributions

Find the top 10 observations that have the highest values for PC1 and
PC2.

```{r}
#find top 10 observations
top_PC1 <- rownames(pca_data)[order(pca_data$PC1, decreasing = TRUE)[1:10]]
top_PC2 <- rownames(pca_data)[order(pca_data$PC2, decreasing = TRUE)[1:10]]
print(top_PC1)
print(top_PC2)
```

# FACTOR ANALYSIS:

We will use the Rent_analysis dataframe that has the variables
`price_per_sqft`, `price_per_bed`, and `price_per_bath`, `price`,
`sqft`, `beds`, and `bath.`

## Determine Optimal Number of Factors

Using fa.parrallel we can determine the number of factors to use in a
factor analysis by comparing the eigenvalues of the data's correlation
matrix with those of randomly generated data. Factors with eigenvalues
greater than those from random data are considered meaningful and should
be retained. Factors with eigenvalues smaller than those from random
data are considered to be noise and are discarded.

```{r}
#similar to the elbow method. can see the drop off is at around 3 or 4
fa.parallel(Rent_analysis, fa = "fa", n.iter = 100, show.legend = TRUE,
            main = "Parallel Analysis Scree Plot")
```

## Perform FA with 3 factors

```{r}
#code to perform and print factor analysis results
factor_analysis <- fa(Rent_analysis, nfactors = 3, rotate = "varimax")
print(factor_analysis)
print(factor_analysis$loadings)
```

## Visualize Factor Analysis Structure

Shows the factors as MR1/2/3. The arrows show the factor loadings, how
much each variable is associated with each factor.

```{r}
fa.diagram(factor_analysis, main = "Factor Analysis Diagram")
```

## Model Interpretation with MLE

The factor loadings tell you how strongly each variable is related to
the factors. A higher absolute value indicates a stronger relationship
between the variable and the factor. For factor 1, baths has the
strongest relationship between the variable and the factor. For factor
2, price is the strongest, and for factor 3, beds is the strongest
relationship.

```{r}
#perform FA and print output
factanal_model <- factanal(Rent_analysis, factors = 3, rotation = "none")
print(factanal_model)
cbind(factanal_model$loadings, factanal_model$uniquenesses)
```

## Find Communalities and Uniquenesses

This function calculates the sum of squared factor loadings for each
variable across all factors. The result, communalities, represents the
total proportion of variance for each variable that is explained by the
factors. Uniqueness is also calculated, which is the amount of variance
that is not explained by the factors. Uniquenesses are calculated as
`1 - communalities`. This value indicates how much of each variable\'s
variance is left unexplained after the factors are extracted.

```{r}
communalities <- rowSums(factor_analysis$loadings^2)
uniquenesses <- factor_analysis$uniquenesses
data.frame(communalities, uniquenesses)
```

## Visualize Factor Loadings

```{r}
#bar plot for each factor 
par(mfrow = c(3, 1))
barplot(factor_analysis$loadings[,1], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 1 Loadings")
barplot(factor_analysis$loadings[,2], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 2 Loadings")
barplot(factor_analysis$loadings[,3], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 3 Loadings")
```

## Evaluate Variance with Sum of Squared Loadings

```{r}
#shows the total explained variance by each factor
SS_loadings <- colSums(factor_analysis$loadings^2)
SS_loadings  
```

-   The first factor explains **2.664598** units of variance in the
    dataset. This is the most important factor and explains the largest
    portion of the total variance.

-   The second factor explains **2.278352** units of variance. This is
    the second most important factor, but it explains slightly less
    variance than the first factor.

-   The third factor explains **1.217245** units of variance. This
    factor explains less variance compared to the first two.

## Visualize Factor Scores Over Time

```{r}
#convert to dataframe for analysis
factor_analysis_df = as.data.frame(factor_analysis$scores)
factor_analysis_df
```

Vizualize factor scores over time for each 3 factor.

```{r}
rent_over_time <- Rent_analysis
rent_over_time$date <- Rent_cleaned$date

factor_scores <- data.frame(date = rent_over_time$date, factor_analysis_df)
factor_scores |>
  gather("factor", "score", -date) |>
  ggplot(aes(x = date, y = score)) + 
  geom_line(linewidth = 1) +
  facet_wrap(~ factor, ncol = 1) +
  theme_bw() +
  labs(title = "Factor Scores Over Time (2000-2018)", x = "", y = "Score")



```

....

# CLUSTERING:

### Make Dataframe

Clean up the data set, make sure there are no NA values or infinite
values. Choose the variables we want.

```{r}
#cleanup data for clustering
Rent_factor_data <- Rent_num |>
  select(price, sqft, beds, baths, price_per_sqft, price_per_bed, price_per_bath)

Rent_factor_data <- Rent_factor_data |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .))) %>%
  drop_na()

Rent_factor_data
```

## Elbow Plot for Price vs Sqft

Use the elbow plot to determine the best number of clusters to use. The
elbow plot helps identify the optimal number of clusters by looking for
the "elbow" in the plot:

```{r}
set.seed(123)

# Calculate WSS for different numbers of clusters
wss <- sapply(1:10, function(k) {
  kmeans(Rent_cleaned[, c("price", "sqft")], centers = k, nstart = 25)$tot.withinss
})

# Create an elbow plot
elbow_plot <- ggplot(data.frame(K = 1:10, WSS = wss), aes(x = K, y = WSS)) +
  geom_line() +
  geom_point() +
  labs(title = "Elbow Plot for Optimal Clusters",
       x = "Number of Clusters",
       y = "Within-cluster Sum of Squares") +
  theme_minimal()

# Print the elbow plot
print(elbow_plot)

```

Based on the elbow plot above, we can see that the most optimal amount
of clusters is 8.

## K-Means Clustering sqft by price:

We will now cluster square foot by price using 8 clusters based on the
above elbow plot.

```{r}
set.seed(123)  
k = 8  

kmeans_result = kmeans(Rent_cleaned[, c("price", "sqft")], centers = k)

Rent_cleaned$cluster = as.factor(kmeans_result$cluster)

ggplot(Rent_cleaned, aes(x = sqft, y = price, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and Square Feet",
       x = "Square Feet",
       y = "Price") +
  theme_minimal()


```

## Hierarchical Clustering with 8 Clusters

Using 8 clusters again, we will now calculate the clusters of square
foot and price, this time with hierarchical clusters. Since the above
plot showed meaningful interpretations using 8 clusters, we will
continue to use 8 clusters for hierarchical and k-mean clusters.

```{r}
Rent_for_clusters = Rent_cleaned[, c("price", "sqft")]
distance_matrix <- dist(Rent_for_clusters)

hclust_result <- hclust(distance_matrix, method = "ward.D2")  

dend <- as.dendrogram(hclust_result)
dend <- color_branches(dend, k = 8)  
plot(dend, main = "Colored Dendrogram for Hierarchical Clustering", xlab = "", ylab = "Height")

hclust_clusters <- cutree(hclust_result, k = 8)
Rent_for_clusters$cluster <- hclust_clusters

Rent_for_clusters %>%
  ggplot(aes(x = sqft, y = price, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Hierarchical Clustering (Price vs. Sqft)", color = "Cluster")

```

## K-means Clustering County and Price

```{r}
library(cluster)

Rent_cleaned2 <- Rent_cleaned %>%
  mutate(county = as.numeric(as.factor(county)))  


data_scaled <- Rent_cleaned2 %>%
  mutate(price = scale(price)) %>%
  select(price, county)  


set.seed(123)
k <- 8 

kmeans_result <- kmeans(data_scaled, centers = k)


Rent_cleaned2$cluster <- as.factor(kmeans_result$cluster)


ggplot(Rent_cleaned2, aes(x = county, y = price, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and county",
       x = "county",
       y = "Price") +
  theme_minimal()
```

## K-means Clustering County and Square Foot

```{r}
library(cluster)

Rent_cleaned2 <- Rent_cleaned %>%
  mutate(county = as.numeric(as.factor(county)))  


data_scaled <- Rent_cleaned2 %>%
  mutate(sqft = scale(sqft)) %>%
  select(sqft, county)  


set.seed(123)
k <- 8 

kmeans_result <- kmeans(data_scaled, centers = k)


Rent_cleaned2$cluster <- as.factor(kmeans_result$cluster)


ggplot(Rent_cleaned2, aes(x = county, y = sqft, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and county",
       x = "county",
       y = "sqft") +
  theme_minimal()

# clustered by scaled sqft and encoded counties
```

## Clustering PC1 and PC2:

We will now perform K-means clustering on the principal components
derived from the PCA and visualize the clustering results using a
scatter plot of the first two principal components.

```{r}
set.seed(123)  
k = 8  

kmeans_result_pca = kmeans(Rent_PCA_df[, c("PC1", "PC2")], centers = k)

Rent_PCA_df$cluster = as.factor(kmeans_result_pca$cluster)

ggplot(Rent_PCA_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by PC1 and PC2",
       x = "PC1",
       y = "PC2") +
  theme_minimal()
```

# Summary Statistics Preparing to Cluster

## Average Price per neighborhood:

```{r}
avg_price_per_nhood = Rent_cleaned %>%
  group_by(nhood) %>%               
  summarise(average_price = mean(price, na.rm = TRUE))  

avg_price_per_nhood

```

## Most and least expensive neighborhoods on average by neighborhood:

```{r}
avg_price_per_nhood[which.max(avg_price_per_nhood$average_price), ]
avg_price_per_nhood[which.min(avg_price_per_nhood$average_price), ]

```

## Average square feet per neighborhood:

```{r}
avg_sqft_per_nhood = Rent_cleaned %>%
  group_by(nhood) %>%               
  summarise(average_sqft = mean(sqft, na.rm = TRUE))  

avg_sqft_per_nhood 
```

## Biggest and Smallest houses by sqft on average:

```{r}
avg_sqft_per_nhood[which.max(avg_sqft_per_nhood$average_sqft), ]
avg_sqft_per_nhood[which.min(avg_sqft_per_nhood$average_sqft), ]
```

# Clustering For Neighborhood Averages

We will use the elbow method again to find the value of k that is ideal
for clustering the data set based on the neighborhood values.

## Elbow Method to determine ideal k for price:

```{r}

avg_scaled_data_price = scale(avg_price_per_nhood$average_price)


wss <- sapply(1:10, function(k) {
  kmeans(avg_scaled_data_price, centers = k, nstart = 25)$tot.withinss 
})

plot(1:10, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal Number of Clusters (price)")

```

## Elbow Method to determine ideal k for square feet:

```{r}
avg_scaled_data_sqft = scale(avg_sqft_per_nhood$average_sqft)


wss <- sapply(1:10, function(k) {
  kmeans(avg_scaled_data_sqft, centers = k, nstart = 25)$tot.withinss 
})

plot(1:10, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal Number of Clusters (sqft)")
```

## Neighborhoods clustered by average price:

```{r}

set.seed(123)

kmeans_result = kmeans(avg_scaled_data_price, centers = 8)

avg_price_per_nhood$cluster = kmeans_result$cluster

print(avg_price_per_nhood)

library(ggplot2)
ggplot(avg_price_per_nhood, aes(x = nhood, y = average_price, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "K-means Clustering of neighborhoods Based on Average Price", x = "nhood", y = "Average Price", color = "Cluster")

```

## Average price per cluster:

```{r}

avg_price_per_cluster = avg_price_per_nhood %>%
  group_by(cluster) %>%
  summarise(average_price_per_cluster = mean(average_price))

print(avg_price_per_cluster)
```

## Neighborhoods clustered by average square feet:

```{r}
set.seed(123)

kmeans_result = kmeans(avg_scaled_data_sqft, centers = 8)

avg_sqft_per_nhood$cluster = kmeans_result$cluster

print(avg_price_per_nhood)

library(ggplot2)
ggplot(avg_sqft_per_nhood, aes(x = nhood, y = average_sqft, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "K-means Clustering of neighborhoods Based on Average sqft", x = "nhood", y = "Average sqft", color = "Cluster")
```

## Average sqft per cluster:

```{r}
avg_sqft_per_cluster = avg_sqft_per_nhood %>%
  group_by(cluster) %>%
  summarise(average_sqft_per_cluster = mean(average_sqft))

print(avg_sqft_per_cluster)
```

```{r}
Rent_cleaned_with_outliers <- Rent |>
  filter(!is.na(title), !is.na(county), !is.na(beds), !is.na(baths), !is.na(sqft))

Rent_cleaned_with_outliers
```

# More Analysis on Clusters

## PAM Clustering:

This is similar to K-means clustering but uses medoids instead of
centroids as the cluster representatives.

```{r}
Rent_cleaned_with_outliers =  Rent_cleaned_with_outliers %>%
  select(price, sqft)

pam_result = pam(Rent_cleaned_with_outliers, k = 3)

print(pam_result$clustering)

print(pam_result$medoids)


Rent_cleaned_with_outliers_df = as.data.frame(Rent_cleaned_with_outliers)
Rent_cleaned_with_outliers_df$cluster = as.factor(pam_result$clustering)


ggplot(Rent_cleaned_with_outliers_df, aes(x = sqft, y = price, color = cluster)) +
  geom_point(size = 1) +
  theme_minimal() +
  labs(title = "PAM Clustering", x = "sqft", y = "price")

```

## Cluster Centers, Profiling, and Heatmap

```{r}
set.seed(123)

kmeans_result_seven = kmeans(Rent_factor_data, centers = 4)

data_scaled_seven$cluster = kmeans_result$cluster
```

These visualizations and summaries allow you to interpret clusters from
your K-means result by exploring the central tendencies (cluster
centers), how variables differentiate between clusters, and which
clusters are most distinct based on their profiles.

```{r}
Rent_factor_data$cluster <- kmeans_result_seven$cluster

# Cluster centers heatmap
cluster_centers <- as.data.frame(kmeans_result_seven$centers)
cluster_centers$cluster <- as.factor(1:nrow(cluster_centers))
cluster_centers_numeric <- cluster_centers[, -ncol(cluster_centers)]
pheatmap(cluster_centers_numeric, main = "Cluster Centers Heatmap")

# Cluster profile
cluster_profile <- Rent_factor_data |>
  group_by(cluster) |>
  summarise(across(c(price, sqft, beds, baths), 
                   list(mean = ~mean(., na.rm = TRUE), 
                        sd = ~sd(., na.rm = TRUE))))
print(cluster_profile)

# Melt and plot heatmap
cluster_centers_melt <- melt(cluster_centers, id.vars = "cluster")
ggplot(cluster_centers_melt, aes(x = cluster, y = variable, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = median(cluster_centers_melt$value)) +
  labs(title = "Heatmap of Cluster Centers") +
  theme_minimal()

```

# Citations

1.  Session 1 and 2 Lab: Data Preprocessing and Visualization Notebook
2.  Session 4 Lab: PCA_CaseStudy_NBA_students Notebook
3.  Session 5 Lab: 2024_FA_CaseStudy_InterestRates_students Notebook
4.  Session 6 Lab: Clustering_WHO_2024_students Notebook
5.  Pennington, Kate (2018). Bay Area Craigslist Rental Housing Posts,
    2000-2018. Retrieved from
    <https://github.com/katepennington/historic_bay_area_craigslist_housing_posts/blob/master/clean_2000_2018.csv.zip>.
