---
title: "Statlearninghw1"
author: "Sam Reade, Ava Exelbirt"
date: "2024-11-08"
output: html_document 
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statistical Learning Homework 1:

```{r libraries}
#install libraries needed 
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggcorrplot")
install.packages("psych")
install.packages("factoextra")
install.packages("pheatmap")
install.packages("dbscan")
library(dbscan)
library(pheatmap)
library(cluster)
library(factoextra)
library(psych)
library(ggcorrplot)
library(dplyr)
library(tidyr)
library(ggplot2)
library(dendextend)
library(reshape2)
library(mclust)
```

# Data Processing Step

Using a data set from TidyTueday that contains information about rent
prices for San Fransisco houses. The data set contains information
regarding date of rental, location of the house (like city and county
for example), information regarding size of house, and price of rent.
The original data set Rent had 200,769 observations among 17 variables.
However, as part of our data preprocessing we chose to drop rows that
contain NA values and drop rows that contain observations with outliers,
as we want our data set to have the most thorough representatife data as
possible. Considering there were still thousands of observations
regarding rental prices in San Francisco, dropping these rows were a
better choice than replacing their values, as we chose to only perform
analysis on real data.

The variables we chose to include in our model were year, neighborhood,
city, county, price, beds, baths, soft, and title. Year is the year the
house was on the market, neighborhood, city, and county outline the area
the house is located in in San Fransisco, price is the rental price is
US dollars per one month of rent, beds is the amount of bedrooms the
house has, baths is the amount of bathrooms the house has, sqft is the
amount of sqft the house has, and title is the title of the rental
listing posting. We chose these variables because they are the most
representative

# Goals

We are choosing to look at

Goals: - predict neighborhood / county by price - predict price by
square footage - find relationships between year and price - predict \#
of beds by price or price by number of beds - maybe do something with
embeddings for the titles for more info about home - correlate with
price

```{r}
Rent_df = tidytuesdayR::tt_load('2022-07-05')
```

```{r}
Rent = Rent_df$rent
Permits = Rent_df$sf_permits
Construction = Rent_df$new_construction
```

# Data Preprocessing:

```{r}
head(Rent)
dim(Rent)
```

# Data Cleanup:

```{r}
# Check number of Duplicate Rows: 
num_duplicates = sum(duplicated(Rent))
print(num_duplicates)
```

# Checking unique values of categorical features:

# Year values:

```{r}
unique_year = unique(as.factor(Rent$year))
print(unique_year)

years_count = table(Rent$year)
print(years_count)
```

# Neighborhood values:

```{r}
unique_nhood = unique(as.factor(Rent$nhood))
print(unique_nhood)

nhood_count = table(Rent$nhood)
print(nhood_count)

# 167 nhoods
```

# City values:

```{r}
unique_city = unique(as.factor(Rent$city))
print(unique_city)

city_count = table(Rent$city)
print(city_count)

# 104 cities
```

# County values:

```{r}
unique_county = unique(as.factor(Rent$county))
print(unique_county)

county_count = table(Rent$county)
print(county_count)

# 10 counties
```

# Removing unnecessary columns:

```{r}
Rent = Rent %>% select(-room_in_apt, -lat, -lon, -address, -descr, -details)
head(Rent)
```

# Checking count of rows with at least 1 NA value:

```{r}
count_na_rows = sum(apply(Rent, 1, function(x) any(is.na(x))))
print(count_na_rows)
```

# Checking count of NA value in each column:

```{r}
na_count_year = sum(is.na(Rent$year))
print(paste("NA count year:", na_count_year))

na_count_nhood = sum(is.na(Rent$nhood))
print(paste("NA count nhood:", na_count_nhood))

na_count_city = sum(is.na(Rent$city))
print(paste("NA count city:", na_count_city))

na_count_county = sum(is.na(Rent$county))
print(paste("NA count county:", na_count_county))

na_count_price = sum(is.na(Rent$price))
print(paste("NA count price:", na_count_price))

na_count_beds = sum(is.na(Rent$beds))
print(paste("NA count beds:", na_count_beds))

na_count_baths = sum(is.na(Rent$baths))
print(paste("NA count baths:", na_count_baths))

na_count_sqft = sum(is.na(Rent$sqft))
print(paste("NA count sqft:", na_count_sqft))

na_count_title = sum(is.na(Rent$title))
print(paste("NA count title:", na_count_title))

```

# Removing rows with NAs for title, county, beds, baths, and sqft

```{r}
Rent_cleaned <- Rent |>
  filter(!is.na(title), !is.na(county), !is.na(beds), !is.na(baths), !is.na(sqft))

Rent_cleaned
```

# Checking for complete rows (no NAs):

```{r}
complete_rows = sum(complete.cases(Rent))

print(complete_rows)
```

# Checking for Outliers:

```{r}

detective = function(column) {

  Q1 = quantile(column, 0.25, na.rm = TRUE)
  Q3 = quantile(column, 0.75, na.rm = TRUE)
  IQR = Q3 - Q1
  
  lower_bound = Q1 - 1.5 * IQR
  upper_bound = Q3 + 1.5 * IQR
  
  outliers = column < lower_bound | column > upper_bound
  
  return(outliers)
}

price_outliers = detective(Rent_cleaned$price)
num_outliers_price = sum(price_outliers)
print(paste("Price outlier count:", num_outliers_price))

bed_outliers = detective(Rent_cleaned$beds)
num_outliers_bed = sum(bed_outliers)
print(paste("Bed outlier count:", num_outliers_bed))

sqft_outliers = detective(Rent_cleaned$sqft)
num_outliers_sqft = sum(sqft_outliers)
print(paste("sqft outlier count:", num_outliers_sqft))

bath_outliers = detective(Rent_cleaned$baths)
num_outliers_bath = sum(bath_outliers)
print(paste("Bath outlier count:", num_outliers_bath))
```

# Visualizing outliers:

```{r}
boxplot(Rent_cleaned$price, 
        main = "Box Plot of Rent Prices", 
        ylab = "Price", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$baths, 
        main = "Box Plot of baths", 
        ylab = "baths", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$sqft, 
        main = "Box Plot of sqft", 
        ylab = "sqft", 
        col = "lightblue",
        outline = TRUE)  

boxplot(Rent_cleaned$beds, 
        main = "Box Plot of beds", 
        ylab = "beds", 
        col = "lightblue",
        outline = TRUE)  


```

# Remove Outliers:

```{r}
Rent_cleaned <- Rent_cleaned |>
  filter(
    !detective(price),
    !detective(beds),
    !detective(baths),
    !detective(sqft)
  )
```

# Correlation Analysis:

```{r}
numeric_complete_rent <- Rent_cleaned |> select(where(is.numeric))

cor_matrix = cor(numeric_complete_rent, use = "pairwise.complete.obs")

ggcorrplot(cor_matrix, lab = TRUE)
```

# Histogram of Price:

```{r}
ggplot(Rent_cleaned, aes(x = price)) +
  geom_histogram(binwidth = 250, color = "black", fill = "skyblue") +
  labs(title = "Histogram of Price", x = "Price", y = "Count") +
  theme_minimal() +
  coord_cartesian(xlim = c(0, 20000))
```

# Histogram of sqft:

```{r}
ggplot(Rent_cleaned, aes(x = sqft)) +
  geom_histogram(binwidth = 100, color = "black", fill = "skyblue") +
  labs(title = "Histogram of Price", x = "sqft", y = "Count") +
  theme_minimal() + 
  coord_cartesian(xlim = c(0, 10000))
```

# Bar Plot of beds:

```{r}
ggplot(Rent_cleaned, aes(x = beds)) +
  geom_bar(color = "black", fill = "skyblue") +
  labs(title = "Bar Plot of beds", x = "Number of beds", y = "Count") +
  theme_minimal()
```

# Bar Plot of baths:

```{r}
ggplot(Rent_cleaned, aes(x = baths)) +
  geom_bar(color = "black", fill = "skyblue") +
  labs(title = "Bar Plot of baths", x = "Number of baths", y = "Count") +
  theme_minimal()
```

# Bar Plot for year:

```{r}
ggplot(Rent_cleaned, aes(x = factor(year))) +  
  geom_bar(color = "black", fill = "skyblue") +  
  labs(title = "Bar Plot of Year", x = "Year", y = "Count") +  
  theme_minimal()
```

-   For nhood beware has some countys in it...

# Feature Engineering:

```{r}
Rent_num = Rent_cleaned

price_per_sqft = Rent_cleaned$price / Rent_cleaned$sqft 
price_per_bed = Rent_cleaned$price / Rent_cleaned$beds
price_per_bath = Rent_cleaned$price / Rent_cleaned$baths

Rent_num$price_per_sqft = price_per_sqft
Rent_num$price_per_bed = price_per_bed
Rent_num$price_per_bath = price_per_bath

Rent_num = Rent_num |> select(-title)

Rent_num
```

# Visualizations of the New Features

```{r}
ggplot(Rent_num, aes(x = price_per_sqft)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightgreen") +
  labs(title = "Histogram of Price per Sqft", x = "Price per Sqft", y = "Count") +
  theme_minimal()

ggplot(Rent_num, aes(x = price_per_bed)) +
  geom_histogram(binwidth = 100, color = "black", fill = "lightcoral") +
  labs(title = "Histogram of Price per Bed", x = "Price per Bed", y = "Count") +
  theme_minimal()

ggplot(Rent_num, aes(x = price_per_bath)) +
  geom_histogram(binwidth = 200, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Price per Bed", x = "Price per Bed", y = "Count") +
  theme_minimal()
```

# EDA Visualizations for Relationships Among Variables:

```{r}
ggplot(Rent_num, aes(x = sqft, y = price)) + 
  geom_point(aes(color = price), alpha = 0.6) + 
  scale_color_gradient(low = "blue", high = "red") + 
  labs(
    title = "Price vs. Square Footage",
    x = "Square Footage",
    y = "Price"
  ) + 
  theme_minimal() + 
  theme(legend.position = "none")


ggplot(Rent_num, aes(x = nhood, y = price)) +
  geom_boxplot(fill = "lightblue", color = "blue", outlier.colour = "red") +
  labs(title = "Price Distribution by Neighborhood", 
       x = "Neighborhood", 
       y = "Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(Rent_num, aes(x = sqft, y = price_per_sqft)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(title = "Price per Square Foot vs. Square Footage",
       x = "Square Footage",
       y = "Price per Square Foot") +
  theme_minimal()


ggplot(Rent_num, aes(x = baths, y = price_per_bath)) +
  geom_point(color = "green", alpha = 0.6) +
  labs(title = "Price per Bath vs. Number of Baths",
       x = "Number of Baths",
       y = "Price per Bath") +
  theme_minimal()
```

# Scale Numeric Features

Standardize/normalize numeric features in dataset.

```{r}
numeric_cols <- Rent_num |> select(where(is.numeric)) |> names()
Rent_num[numeric_cols] <- scale(Rent_num[numeric_cols])
```

# Correlation Analysis with the new features:

```{r}
numeric_complete_rent2 <- Rent_num |> select(where(is.numeric))

cor_matrix2 <- cor(numeric_complete_rent2, use = "pairwise.complete.obs")

ggcorrplot(cor_matrix2, lab = TRUE)
```

# Data Set/Extra Cleanup for PCA:

***For predict neighborhood / county / city by price:\
Rent_price (190,618 x 8) --\> id, date, year, nhood, city, county,
price, title***

***For predicting price by square footage: Rent_cleaned_sqft (61,056 x
9) --\> id, date, year, nhood, city, county, sqft, price, title***

***For finding relationships between year and price: Rent_price (190,618
x 8) --\> id, date, year, nhood, city, county, price, title***

***For predicting \# of beds by price: Rent_price (190,618 x 8) --\> id,
date, year, nhood, city, county, price, title***

```{r}
Rent_analysis <- Rent_num |>
  select(price, sqft, beds, baths, price_per_sqft, price_per_bed, price_per_bath, year, date)

# Step 2: Clean the data by handling NA and infinite values
# Replace NA values with the column mean, and handle Inf values
Rent_analysis <- Rent_analysis |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>
  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .))) %>%
  drop_na() 
```

# PCA ANALYSIS:

***NEED TO CHANGE WRITING BELOW:***

***Engineer new features: - price_per_sqft = price / sqft --\>
"Rent_cleaned_sqft" - price_per_bed = price / bed --\> "Rent_price" -
price_per_bathroom = price / bath --\> "Rent_cleaned_baths"***

***7 features to work with***

***DF: Rent_for_PCA***

***CHANGE WRIITNG ABOVE:***

## Perform PCA

First standardize the data, then perform PCA, and view the PCA loadings.
The PCA loadings (the relationship between original variables and
principal components). This will help understand how each feature
contributes to each principal component

```{r PCA}
Rent_PCA <- Rent_analysis |>
  select(where(~ !all(is.na(.))))

Rent_PCA <- Rent_PCA |>
  select(where(~ var(.) != 0))

Rent_PCA_scaled <- scale(Rent_PCA)

pca_result <- prcomp(Rent_PCA_scaled, center = TRUE, scale. = TRUE)

# Scree Plot using factoextra
fviz_screeplot(pca_result, addlabels = TRUE, title = "Scree Plot")

summary(pca_result)

print(pca_result$rotation)
```

## Proportion of Variance for each PC

This tells you how much of the variability in the data is explained by
each component

```{r}
proportion_variance <- summary(pca_result)$importance[2, ]
print(proportion_variance)
```

## Visualize the first two PCs.

```{r}
Rent_PCA_df = data.frame(PC1 = pca_result$x[, 1], PC2 = pca_result$x[, 2])
Rent_PCA_df
```

```{r}
ggplot(Rent_PCA_df, aes(x = PC1, y = PC2)) +
  geom_point(color = "blue", alpha = 0.6) +  
  theme_minimal() +
  labs(title = "PCA: First vs. Second Principal Component", 
       x = "Principal Component 1", y = "Principal Component 2")


barplot(pca_result$rotation[,1], las = 2, col = "darkblue", main = "Loadings for PC1")

barplot(pca_result$rotation[,2], las = 2, col = "darkblue", main = "Loadings for PC2")
```

## Visualize the Variance of PCs.

```{r}
variance_explained <- summary(pca_result)$importance[2, ]
ggplot(data.frame(PC = 1:length(variance_explained), Variance_Explained = variance_explained), 
       aes(x = factor(PC), y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Variance Explained by Each Principal Component", x = "Principal Component", y = "Variance Explained")
```

## Normalized

```{r}
sum(pca_result$rotation[,1]^2)
```

## Contribution of Variables to PCs

```{r}
fviz_contrib(pca_result, choice = "var", axes = 1, title = "Contribution of Variables to PC1")
fviz_contrib(pca_result, choice = "var", axes = 2, title = "Contribution of Variables to PC2")

```

## PCA Bipolot

```{r}
fviz_pca_biplot(pca_result, repel = TRUE, title = "PCA Biplot")
                
```

## Create dataframe with results

The principal components (scores) are stored in pca_result\$x, which we
convert into a dataframe.

```{r}
pca_data <- data.frame(pca_result$x) 
```

## Individual Component Contributions

```{r}
top_PC1 <- rownames(pca_data)[order(pca_data$PC1, decreasing = TRUE)[1:10]]
top_PC2 <- rownames(pca_data)[order(pca_data$PC2, decreasing = TRUE)[1:10]]
print(top_PC1)
print(top_PC2)
```

# Factor Analysis:

Select relevant columns for factor analysis, only numeric ones:
`price_per_sqft`, `price_per_bed`, and `price_per_bath`, `price`,
`sqft`, `beds`, and `bath.` We will use the Rent_for_PCA dataframe.

## Determine Optimal Number of Factors

```{r}
fa.parallel(Rent_analysis, fa = "fa", n.iter = 100, show.legend = TRUE,
            main = "Parallel Analysis Scree Plot")
```

## Perform FA with 3 factors

```{r}
factor_analysis <- fa(Rent_analysis, nfactors = 3, rotate = "varimax")
print(factor_analysis)
print(factor_analysis$loadings)
```

## Visualize Factor Analysis Structure

```{r}
fa.diagram(factor_analysis, main = "Factor Analysis Diagram")
```

## Model Interpretation with MLE

```{r}
factanal_model <- factanal(Rent_analysis, factors = 3, rotation = "none")
print(factanal_model)
cbind(factanal_model$loadings, factanal_model$uniquenesses)
```

## Find Communalities and Uniquenesses

```{r}
communalities <- rowSums(factor_analysis$loadings^2)
uniquenesses <- factor_analysis$uniquenesses
data.frame(communalities, uniquenesses)
```

## Visualize Factor Loadings

```{r}
par(mfrow = c(3, 1))
barplot(factor_analysis$loadings[,1], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 1 Loadings")
barplot(factor_analysis$loadings[,2], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 2 Loadings")
barplot(factor_analysis$loadings[,3], las = 2, col = "darkblue", ylim = c(-1, 1), main = "Factor 3 Loadings")
```

## Evaluate Variance with Sum of Squared Loadings

```{r}
SS_loadings <- colSums(factor_analysis$loadings^2)
SS_loadings  
```

## Visualize Factor Scores

```{r}
factor_analysis_df = as.data.frame(factor_analysis$scores)
factor_analysis_df
```

```{r}
factor_scores <- data.frame(date = Rent_analysis$date, factor_analysis_df)
library(ggplot2)
factor_scores |>
  gather("factor", "score", -date) |>
  ggplot(aes(x = date, y = score)) + 
  geom_line(linewidth = 1) +
  facet_wrap(~ factor, ncol = 1) +
  theme_bw() +
  labs(title = "Factor Scores Over Time (2000-2018)", x = "", y = "Score")
```

## Oblimin Rotation for Correlated Factors

```{r}
factor_analysis_oblimin <- fa(Rent_analysis, nfactors = 3, rotate = "oblimin")
print(factor_analysis_oblimin$loadings)

#Check communalities for potential issues
communalities <- rowSums(factor_analysis_oblimin$loadings^2)
print(communalities)

# Not sure what the issue is with this but need to be mindful it says they are probably incorrect
```

## Alternate Methods

```{r}
# Alternative rotation if needed
factor_analysis_promax <- fa(Rent_analysis, nfactors = 3, rotate = "promax")
print(factor_analysis_promax$loadings)

#Principal Axis Factoring (PAF)
factor_analysis_paf <- fa(Rent_analysis, nfactors = 3, fm = "pa")
print(factor_analysis_paf$loadings)

#Weighted Least Squares (WLS)
factor_analysis_wls <- fa(Rent_analysis, nfactors = 3, fm = "wls")
print(factor_analysis_wls$loadings)

#Generalized Least Squares (GLS)
factor_analysis_gls <- fa(Rent_analysis, nfactors = 3, fm = "gls")
print(factor_analysis_gls$loadings)

#Alpha Factoring 
factor_analysis_alpha <- fa(Rent_analysis, nfactors = 3, fm = "alpha")
print(factor_analysis_alpha$loadings)
```

....

# CLUSTERING

Property Segmentation: Cluster sqft, price, beds, and baths to group
similar types of properties --\> clusters might represent luxury homes,
small apartments, or family homes\
- complete_Rent

Market Segmentation by Location: Cluster based on city, county, or nhood
combined with property attributes like price and sqft to find
location-based patterns - complete_rent

## K-Means Clustering with 3 clusters

### Perform Clustering + Visualization

```{r}
Rent_factor_data <- Rent_num %>%
  select(price, sqft, beds, baths, price_per_sqft, price_per_bed, price_per_bath)

Rent_factor_data <- Rent_factor_data %>%
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .))) %>%
  drop_na()

Rent_factor_data
```

```{r}
Rent_factor_data_scaled <- scale(Rent_factor_data)

set.seed(123)  
kmeans_result <- kmeans(Rent_factor_data_scaled, centers = 3, nstart = 25) 
print(kmeans_result)  

Rent_factor_data$cluster <- kmeans_result$cluster

Rent_factor_data |>
  ggplot(aes(x = price, y = sqft, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "K-means Clustering (Price vs. Sqft)", color = "Cluster")
```

### Silhouette Analysis for Optimal Clusters

```{r}
silhouette_score <- silhouette(kmeans_result$cluster, dist(Rent_factor_data_scaled))
plot(silhouette_score, main = "Silhouette Plot for K-means Clustering")
```

### Cluster Centers, Profiling, and Heatmap

```{r}
cluster_centers <- as.data.frame(kmeans_result$centers)
pheatmap(cluster_centers, main = "Cluster Centers Heatmap")

cluster_profile <- Rent_factor_data %>%
  group_by(cluster) %>%
  summarise(across(c(price, sqft, beds, baths), list(mean = mean, sd = sd), na.rm = TRUE))
print(cluster_profile)

cluster_centers_melt <- melt(cluster_centers, id.vars = "cluster")

ggplot(cluster_centers_melt, aes(x = cluster, y = variable, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = median(cluster_centers_melt$value)) +
  labs(title = "Heatmap of Cluster Centers") +
  theme_minimal()

```

## Hierarchical Clustering with 3 Clusters

```{r}
Rent_factor_data_scaled <- scale(Rent_factor_data)
distance_matrix <- dist(Rent_factor_data_scaled)

hclust_result <- hclust(distance_matrix, method = "ward.D2")  

dend <- as.dendrogram(hclust_result)
dend <- color_branches(dend, k = 3)  
plot(dend, main = "Colored Dendrogram for Hierarchical Clustering", xlab = "", ylab = "Height")

hclust_clusters <- cutree(hclust_result, k = 3)
Rent_factor_data$cluster <- hclust_clusters

Rent_factor_data %>%
  ggplot(aes(x = price, y = sqft, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Hierarchical Clustering (Price vs. Sqft)", color = "Cluster")

```

# K-means Clustering (trying to find meaningful clusters)

-   high rent areas (cities)

```{r}
library(cluster)

Rent_cleaned <- Rent_cleaned %>%
  mutate(city = as.numeric(as.factor(city)))  


data_scaled <- Rent_cleaned %>%
  mutate(price = scale(price)) %>%
  select(price, city)  


set.seed(123)
k <- 8 

kmeans_result <- kmeans(data_scaled, centers = k)


Rent_cleaned$cluster <- as.factor(kmeans_result$cluster)


ggplot(Rent_cleaned, aes(x = city, y = price, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and City",
       x = "City",
       y = "Price") +
  theme_minimal()

# clustered by scaled price and encoded cities
```

# Clustering sqft by price:

```{r}
complete_Rent = Rent[complete.cases(Rent), ]
complete_Rent

set.seed(123)  
k = 10  

kmeans_result = kmeans(complete_Rent[, c("price", "sqft")], centers = k)

complete_Rent$cluster = as.factor(kmeans_result$cluster)

ggplot(complete_Rent, aes(x = sqft, y = price, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and Square Feet",
       x = "Square Feet",
       y = "Price") +
  theme_minimal()


```

# Clustering Price by sqft:

```{r}
set.seed(123)  
k = 10  

kmeans_result = kmeans(complete_Rent[, c("price", "sqft")], centers = k)

complete_Rent$cluster = as.factor(kmeans_result$cluster)

ggplot(complete_Rent, aes(x = price, y = sqft, color = cluster)) +
  geom_point(size = 1) +
  labs(title = "Rent Clusters by Price and Square Feet",
       x = "Square Feet",
       y = "Price") +
  theme_minimal()
```

# Average Price per neighborhood:

```{r}
avg_price_per_nhood = Rent_cleaned %>%
  group_by(nhood) %>%               
  summarise(average_price = mean(price, na.rm = TRUE))  

avg_price_per_nhood

```

# Average square feet per neighborhood:

```{r}
avg_sqft_per_nhood = Rent_cleaned %>%
  group_by(nhood) %>%               
  summarise(average_sqft = mean(sqft, na.rm = TRUE))  

avg_sqft_per_nhood 
```

# Elbow Method to determine ideal k for price:

```{r}

avg_scaled_data_price = scale(avg_price_per_nhood$average_price)


wss <- sapply(1:10, function(k) {
  kmeans(avg_scaled_data_price, centers = k, nstart = 25)$tot.withinss 
})

plot(1:10, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal Number of Clusters (price)")

```

# Elbow Method to determine ideal k for square feet:

```{r}
avg_scaled_data_sqft = scale(avg_sqft_per_nhood$average_sqft)


wss <- sapply(1:10, function(k) {
  kmeans(avg_scaled_data_sqft, centers = k, nstart = 25)$tot.withinss 
})

plot(1:10, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of Clusters (k)", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal Number of Clusters (sqft)")
```

# Neighborhoods clustered by average price:

```{r}

set.seed(123)

kmeans_result = kmeans(avg_scaled_data_price, centers = 8)

avg_price_per_nhood$cluster = kmeans_result$cluster

print(avg_price_per_nhood)

library(ggplot2)
ggplot(avg_price_per_nhood, aes(x = nhood, y = average_price, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "K-means Clustering of neighborhoods Based on Average Price", x = "City", y = "Average Price", color = "Cluster")

```

# Neighborhoods clustered by average square feet:

```{r}
set.seed(123)

kmeans_result = kmeans(avg_scaled_data_sqft, centers = 8)

avg_sqft_per_nhood$cluster = kmeans_result$cluster

print(avg_price_per_nhood)

library(ggplot2)
ggplot(avg_sqft_per_nhood, aes(x = nhood, y = average_sqft, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "K-means Clustering of neighborhoods Based on Average sqft", x = "City", y = "Average sqft", color = "Cluster")
```

## Other Possible Methods

## Density Based Spatial Clustering of Applications with Noise (DBSCAN)

```{r}
data_scaled <- Rent_cleaned |>
  mutate(price = scale(price), beds = scale(beds)) |>
  select(price, beds)

dbscan_result <- dbscan(data_scaled, eps = 0.5, minPts = 5)

Rent_cleaned$dbscan_cluster <- as.factor(dbscan_result$cluster)

ggplot(Rent_cleaned, aes(x = beds, y = price, color = dbscan_cluster)) +
  geom_point(size = 1) +
  labs(title = "DBSCAN Clustering of Rent Prices and Square Feet", color = "Cluster") +
  theme_minimal()

```

## Gaussian Mixture Models (GMM)

```{r}
gmm_result <- Mclust(data_scaled)

Rent_cleaned$gmm_cluster <- as.factor(gmm_result$classification)

ggplot(Rent_cleaned, aes(x = beds, y = price, color = gmm_cluster)) +
  geom_point(size = 1) +
  labs(title = "GMM Clustering of Rent Prices and Beds", color = "Cluster") +
  theme_minimal()
```

# Citations

1.  Session 1 and 2 Lab: Data Preprocessing and Visualization Notebook
2.  Session 4 Lab: PCA_CaseStudy_NBA_students Notebook
3.  Session 5 Lab: 2024_FA_CaseStudy_InterestRates_students Notebook
4.  Session 6 Lab: Clustering_WHO_2024_students Notebook
5.  Pennington, Kate (2018). Bay Area Craigslist Rental Housing Posts,
    2000-2018. Retrieved from
    <https://github.com/katepennington/historic_bay_area_craigslist_housing_posts/blob/master/clean_2000_2018.csv.zip>.
